\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-05-11 papers/paper1/ms.tex

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
This method can only be used with one-dimensional datasets and the covariance
function must be represented by a mixture of exponentials, we will return to a
discussion of what this means in detail in the following sections.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
More precisely, the matrix can be embedded in an equivalent higher-dimensional
system that can be solved efficiently using a sparse or banded solver.
As we will discuss in the following pages, this choice of kernel function is
not as restrictive as it first seems because it arises naturally in some
physical systems and it can be used as an effective model in many other cases.
This method is especially appealing compared to other similar methods~--~we
return to these below~--~because it is exact, flexible, simple, and fast.

@@

In the following pages, we motivate the general problem of GP modeling,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released well-tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo}
\citep{Foreman-Mackey:2017}.

>>>>

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
\response{This method is a generalization of a method developed by
\citet{Ambikasaran:2015} 
 ({\bf This is no longer correct -different Ambikasaran result!})}
that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
This method can only be used with one-dimensional datasets and the covariance
function must be represented by a mixture of exponentials; we will return to a
discussion of what this means in detail in the following sections.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\response{More precisely, the matrix can be embedded in an equivalent higher-dimensional
system that can be solved efficiently using a sparse or banded solver. Replace
with(?): The matrix can be expressed as the sum of semi-separable matrices, each
of which is constructed from the outer product of one-dimensional vectors.  These
are a type of low-rank matrix, which may be solved by Cholesky decomposition
using a novel technique developed by one of us (SA).}
As we will discuss in the following pages, this choice of kernel function is
not as restrictive as it first seems because it arises naturally in some
physical systems and it can be used as an effective model in many other cases.
This method is especially appealing compared to other similar methods~--~we
return to these below~--~because it is exact, flexible, simple, and fast.

@@

In the following pages, we motivate the general problem of GP modeling,
describe the \response{(delete) previously published} scalable method 
\citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate 
the model's application on various real and simulated data sets.
Alongside this paper, we have released well-tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo}
\citep{Foreman-Mackey:2017}.
